% !TEX root = ../main.tex

\chapter{Autonomous Driving Systems}
\label{chapter:background}

%* fix the phrasing and continue to expand the current idea
This chapter serves to present a more detailed and technical description of previous approaches. The focus will be on their shortcomings to provide a more universal yet safe implementation
of Autonomous Navigational Systems.

\section{Modular Stack}
\label{section:ModularStack}
%* fix and check all citations

Since the early 2000s, researchers and industry manufacturers have adopted the modular stack as the conventional architecture for autonomous vehicles \cite{jain_Autonomy_20_2021, wang_VersatileEfficientReinforcement_2022,tampuu_SurveyEndtoEndDriving_2022}.
The Defense Advanced Research Projects Agency (DARPA) Grand Challenges of 2004 and 2005 accelerated mainstream adoption of the modular stack when teams demonstrated that the modular pipeline could autonomously navigate a demanding off-road desert course within a prescribed time limit \cite{tampuu_SurveyEndtoEndDriving_2022,wang_VersatileEfficientReinforcement_2022,thrun_stanley_2006}.
Engineers described the challenge as solving a software problem, defining deterministic rules and interfaces for each step of the cohesive pipeline \cite{thrun_stanley_2006, montemerlo_winning_2006}.

A robot named Stanley became the first unmanned vehicle to traverse the 212\,km course at the 2005 DARPA Grand Challenge \cite{tampuu_SurveyEndtoEndDriving_2022, hoffmann_AutonomousAutomobileTrajectory_2007, thrun_stanley_2006}.
Stanley's software pipeline is described as comprising of approximately 30 modules organized into six functional layers \cite{tampuu_SurveyEndtoEndDriving_2022, thrun_stanley_2006, montemerlo_winning_2006}.
Fusing input data from various cameras, laser range finders, and other positional sensors, the pipeline routed this stream through the perception layer to construct the sorrounding environment and the vehicle's orientation within it \cite{thrun_stanley_2006}.
%? There was no global path planning involved (change: the role of Stanley’s path planner is primarily local obstacle avoidance.)
Achieving high-speed driving yet careful local obstacle avoidance through extensive tweaking: refining software components iteratively, and tuning modules in isolation until the full vehicle exhibited the desired behavior \cite{thrun_stanley_2006, montemerlo_winning_2006}.
This module based development strategy offers clear interpretability and allows engineers to optimize individual units against well-defined inputs and outputs.
%? Talk about more recent benefit of modular stack check Jeremy paper

However, The DARPA Grand Challenges additionally illustrated the limitations of this approach.
In the 2004 challenge, none of the competing vehicles completed more than 5\% of the course through the Mojave Desert \cite{ogretmen_local_2025, thrun_stanley_2006, montemerlo_winning_2006}.
One year later, only five of the 23 entrants crossed the finish line, indicating that reliable completion of the course still required complex and meticulous tuning across the entire software pipeline \cite{tampuu_SurveyEndtoEndDriving_2022,wang_VersatileEfficientReinforcement_2022,jain_Autonomy_20_2021,kocic_end-to-end_2019,thrun_stanley_2006}.
This narrow margin of success is indicative of the error propagation within the modular stack, where modeling approximations and estimation uncertainties accumulate downstream and compound to degrade planning and control performance \cite{jain_Autonomy_20_2021,tampuu_SurveyEndtoEndDriving_2022,wang_VersatileEfficientReinforcement_2022,kocic_end-to-end_2019}.

% \cite{gonzalez_ReviewMotionPlanning_2016,hoffmann_AutonomousAutomobileTrajectory_2007,ogretmen_local_2025,werling_OptimalTrajectoryGeneration_2010,rosero_integrating_2024,thrun_stanley_2006,urmson_autonomous_2008}
%? More recently, hybrid architectures have begun to integrate modular perception and control pipelines with learned planning components, combining the debuggability of the classical stack with data-driven adaptability \cite{rosero_integrating_2024}.

\section{End-to-End Neural Network Stack}
\label{section:NNStack}

% Include specific examples of use cases. Discuss further about global optimal point using gradient descent methods.
% There are some variations to this approach with some choosing to cut off the optimization task at the planning task which produces different driving modalities: including creating
% waypoints to pass on later, etc .

In the plain end-to-end neural network stack, the entire pipeline is treated as a single learned task with the goal of predicting what the car's actuators should do. 
Most commonly 

\cite{hu_review_2026,tampuu_SurveyEndtoEndDriving_2022,wang_VersatileEfficientReinforcement_2022,kocic_end-to-end_2019,lee_autonomous_2020,hu_learning_2021,chib_recent_2023}

%* describe IL here mathematically, introduce the concept, what does it aim to do, what are some examples of used IL inside this
Imitation Learning, a machine learning paradigm, uses prerecorded human driving data to perform supervised learning on the driving model.
The model slowly learns to mimic human driving behavior. However, this technique suffers from distribution shift, data bias, and causal confusion \cite{tampuu_SurveyEndtoEndDriving_2022,wang_VersatileEfficientReinforcement_2022}.
These problems come from a mix of having to gather so much data, ensure that your data is diverse enough, and all of that you still cannot be sure that you will cover all scenarios.
Especially when it comes to racing.

%* describe RL here mathematically, introduce the concept, what does it aim to do, what are some examples of used RL inside this
Reinforcement Learning is also a common choice, where the system has more ability to explore and interact with its environment.
Through this continuous interaction with the environment it learns the most optimal behavior based on a policy rewarding system.
For this reason, RL has incentives to explore unorthodox driving styles and behaviors that are rarer to see in real life but yet still physically possible.
No labeled data needed or required so it is easier to set up but also suffers more from data efficiency than IL. also harder to train with high dim data \cite{wang_VersatileEfficientReinforcement_2022}

Reinforcement Learning (RL) is a promising approach to address the distribution shift problem.
It aims to maximize cumulative rewards over time by interacting with the environment, and the network makes driving decisions to obtain rewards or penalties based on its actions.
IL cannot handle novel situations significantly different from the training dataset, RL is robust to this issue as it explores scenario under given environment \cite{chib_recent_2023}.

Both are prone to interpratible and unsafe behavior


\section{Modular Differentiable Stack}
\label{section:ModDiffStack}

% Speak more on the main concept of modular and differentiable stacks 

% Describe work of DiffStack, and how it operates

% Describe DIPP and how it operates

% Conversely, the modular architecture has matured immensely since that time, where improvements to the system are now introducing neural networks for specific tasks.
% The perception task can highly benefit from Convolution Neural Networks (CNNs) utilizing their ability to extract visual information from images and interpret them.
% The prediction task can also be enhanced with the introduction of a Neural Network which outputs the most probable scenario.
% However, these only achieve partial, unit specific, optimizations rather than a full system optimal point.

% Give examples on this specific topic which includes DiffStack, DIPP, etc.

% What have people used previously
%? Focusing on partial combination of the architectures.
The modular architecture has matured over the years with advancements to the pipeline which utilizes NN to handle some of the tasks individually.
This comes after the improvements of NN and specifically Convolution Neural Networks (CNNs) enabling extraction of information from images and interpreting them.
Hence, perception task benefit from using CNNs to read LiDAR or camera data from the car to interpret the vehicle's surroundings.
Furthermore, the prediction task can be enhanced by feeding the data into a NN which can output the most probable scenario to occur in the next time steps. 
Both of these tasks are now no longer hand tuned by the engineers.
However, this approach only achieves partial, unit specific optimizations rather than finding a full system optimization.  

\subsection{DiffStack}
\label{subsection:DiffStack}
There are varying degrees of hybridization. DiffStack explanation which takes in 

1. Prediction is trained for ego planning/control, not generic forecasting. DiffStack reports up to ~15\% higher prediction effectiveness for planning and fewer errors that actually hurt the plan.

2. Substantial gains in open-loop and closed-loop planning metrics compared with separately trained modules.

Joint training beats local training — End-to-end training outperforms training prediction in isolation, even when using handcrafted “planning-aware” auxiliary losses.

Keeps modularity — Still uses explicit prediction, planning, and control blocks; you get co-optimization without a black-box end-to-end network.

\subsection{DIPP}
\label{subsection:DIPP}

Fixes the prediction-planning split — Explicitly targets the two classic modular failures: isolated prediction/planning and hand-tuned planner costs.
Joint training beats sequential training — In both open-loop and closed-loop tests, integrated training outperforms using a separately trained prediction module — on both planning and prediction metrics.
Planning-centric prediction — Predictions are shaped so the planner produces trajectories closer to human driving, not just lower raw forecast error.
Learned cost function — Planner weights are learned from data instead of manually tuned, improving stability and performance (ablations show learnable components matter).

\subsection{MIND}
\label{subsection:MIND}

\chapter{Motion Planner}
\label{chapter:MotionPlanner}

% Talk about different motion planning techniques. Evolution of motion planning in general to how we reached this current planning task.
% Covering the mathematical background to understand how the sampling planner works.
% Discussion on local planner? differentiable local planner?
% Current reach of stack? what can be achieved now with all the other work having been done?

%* Talk about what a motion planner is, what it does specifically. Why the focus on differentiability.

\section{Global vs Local Planning}
\label{section:GlobalvsLocal}

a general overview of global vs local planning. Include definition of global planning as a general big picture plan that the car must follow.
In terms of racing similar to a track where a raceline is known, have good guess what to do. This is typically done offline, achieving a specific goal in mind.

local planning is the car reacting in real time what it needs. obstacle detection, opponent in front or behind, changed racing conditions. Adapt as needed.
Short horizon -> what is the best trajectory

\section{Different Motion Planners}
\label{section:MotionPlanner}

\subsection{Graph and Search Planners}
\label{subsection:GraphandSearch}

This includes A* approaches where they discretize space and then turning them differentiable is difficult. find citations

\subsection{Sampling-Based Planners}
\label{subsection:SamplingBased}

RRT trees and RRT* trees random explore, also hard to differentiate at any point.

\subsection{Frenet Sampling}
\label{subsection:FrenetSampling}

polynomial samplings

\section{Frenet Frame}
\label{section:FrenetFrame}

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}
        % Track
        \draw[thick, double=white, double distance=1.5cm] 
            (0, 0) -- (0, 2) arc (180:90:2) -- (5.5, 4);
        % racing line
        \draw[very thick, dashed, red] 
            plot [smooth, tension=0.6] coordinates {(-0.5, 0) (-0.3, 1.2) (0.9, 3.1) (3.5, 4) (5.5, 4.2)};
        % car
        \fill[black] (0.05,2) circle (2pt);
        \draw[->, thick] (1,0) -- (1.5,0) node[right] {$x$};
        \draw[->, thick] (1,0) -- (1,0.5) node[above] {$y$};
    \end{tikzpicture}
    \caption{Track Representation}
    \label{fig:trackrep}
\end{figure}

The Frenet frame of reference is often used when describing a moving particle along a differentiable curve. The use of Frenet coordinate system
$(s, n)$ allows simpler interpretation of curved paths than using a rigid system such as the Cartesian coordinates $(x, y)$.
The curvilinear Frenet frame consists of the longitudinal path $s$ along the curve and its deviation from said curve $n$ or sometimes called $d$.
% modify the phrasing
An additional reason Frenet is beneficial to use is because it simplifies the process of defining planning path constraints or target goals for the vehicle by separating longitudinal and lateral path directions. 
The curve vehicles follow is called a raceline; it is a mathematically optimal line aiming to achieve a certain goal, most commonly fastest time to complete a full lap
around a circuit, or minimum curvature of the raceline. The raceline may be represented as continuous flat curve:

\begin{equation}
    \mathcal{R} = \{ \textbf{r}_{\text{raceline}}(s) \in \mathbb{R}^2 \ | \ s \in [0, s_{\text{lap}}]\},
\end{equation}
where $s$ is the arc-length distance along the raceline curve, and $s_{\text{lap}}$ the total length of a single lap around the circuit. 
The position vector of the reference curve can be described in simple Cartesian coordinates, for each point $s$ on the curve,
\begin{equation}
    \textbf{r}(s) = \begin{bmatrix}
        x_{\text{raceline}} (s) \\ y_{\text{raceline}} (s)
    \end{bmatrix}.
\end{equation}
Additionally, the tangent vector of the raceline defined as the derivative of the position vector with respect to $s$, and the normal vector is orthogonal to the tangent vector, can both be mathematically described as,
\begin{equation}
    \textbf{t}(s) = \frac{d \textbf{r}(s)}{d s} = \begin{bmatrix} \cos (  \psi_{\text{raceline}}(s)) \\ \sin (\psi_{\text{raceline}}(s)) \end{bmatrix}, \qquad
    \textbf{n}(s) = \begin{bmatrix} - \sin (\psi_{\text{raceline}}(s)) \\ \cos (psi_{\text{raceline}}(s)) \end{bmatrix}.
\end{equation}
Where $\psi (s)$ is the heading angle, relative to the global Cartesian frame, of the raceline at point $s$.

% A sampling planner is currently being used to find possible trajectories for the vehicle to take, a cost function assess all the possible trajectories
% and gives them a score depending on the path's "correctness". The path with the smallest cost is chosen and is then fed to the MPC controller as the ideal
% required trajectory.

\chapter{DiffPlan}
\label{chapter:DiffPlan}

%* What does our specific motion planner do?
%* How many parts is it made out of? 
%* How can we describe it

\section{Sampling Technique}
\label{section:SamplingTechnique}

The original sampling strategy for the motion planner:

Starting with an initial state of the longitudinal and lateral position and velocity of the vehicle, $S_0 = [s_0, \dot{s}_0, n_0, \dot{n}_0]$, respectively.
A set of sample trajectories are generated constrained by the initial longitudinal velocity and lateral position of the vehicle, the samples terminate at
the end states, $S_{\text{end}, ij} = [\dot{s}_{\text{end}, i}, n_{\text{end}, j}, \dot{n}_{\text{end}, j}]$ with a specific candidate ${ij}$. The sampling
strategy is split into first the longitudinal sample component followed by the lateral sample component. 

[WARNING CHANGE LATER] % probably flip the entire description start with what we want to end with. s(t) and the big S matrix 

The longitudinal sampling is initialized by discretizing the time horizon $T$ into equally spaced steps, which results in a discrete time vector. 
Subsequently, the range of permissible longitudinal velocities $[\dot{s}_{\text{min}}, \dot{s}_{\text{max}}]$ at the trajectory's endpoint is determined, 
constrained by the initial velocity, the target velocity, and the velocity derived from the global race line. 
The upper bound of this range is computed as
\begin{equation}
    \dot{s}_{\text{max}} = \text{min}(V_{\text{max}}, \text{max}(\dot{s}_0, \dot{s}_{\text{incr}}) \cdot K_s ),
\end{equation}
where K is a a scaling factor that accounts for a necessary deviation from the raceline, while the velocity increment
\begin{equation}
    \dot{s}_{\text{incr}} = \text{min}(\dot{s}_0 + \Delta \dot{s}, \dot{s}_{rl}(T)),
\end{equation}
This formulation ensures that the velocity range appropriately considers the initial vehicle states, the velocity differences, and the reference trajectory.
The velocity samples $ \{ \dot{s}_{e, i} \}_{i=0}^{N_s}$ are generated within the range, where $N_s$ denotes the number of samples. 
Each sampled end velocity is utilized to construct a longitudinal trajectory using a quadratic polynomial,
\begin{equation}
    s(t) = c_0 + c_1 t + c_2 t^2 ,
\end{equation}
and the corresponding longitudinal velocity is given by,
\begin{equation}
    \dot{s}(t) = c_1 + 2c_2 t.
\end{equation}
The coefficient $c_0, c_1, c_2$ of the polynomial are determined by solving the linear system $A_s c_s = b_s$ with
\begin{equation}
    A_s = \begin{bmatrix}
        1 & 0 & 0 \\
        0 & 1 & 0 \\
        0 & 1 & 2T
    \end{bmatrix}, \quad
    b_s = \begin{bmatrix}
        s_0 \\
        \dot{s}_0\\
        \dot{s}_i
    \end{bmatrix}.
\end{equation}
Where it is constrained by 3.18 and 3.19 as the initial and boundary conditions $s_0, \dot{s}_0, \dot{s}_{e,i}$.
These coefficients ensure that the resulting polynomial maintains continuity with the initial position $s_0$ with the initial velocity $\dot{s}_0$ and a smooth transition to the sampled velocity.
The resulting set of sampled longitudinal trajectories can be represented as a matrix $S_j \in \mathbb{R}^{N_s \times N_T \times 4}$, where each row corresponds to a distinct trajectory profile $S_i(t)$ over the time interval $[0, T]$.
As the final velocities are particularly relevant in racing scenarios, we do not sample the end positions longitudinally [5]; instead, we adjust them according to the polynomial formulation in Eq. 3.18.

For lateral sampling, a cubic polynomial is used to describe the lateral trajectory:
\begin{equation}
    n(t) = c_0 + c_1 t + c_2 t^2 + c_3 t^3,
\end{equation}
with the corresponding lateral velocity $\dot{n}(t)$ is the first derivative of the position function. 
The polynomial coefficients $c_n = [c_0, c_1, c_2, c_3]^T$ are calculated by solving linear equations $A_n c_n = b_n$, where the coefficient matrix $A_n$ defined as
\begin{equation}
    A_n = \begin{bmatrix}
        1 & 0 & 0 & 0\\
        0 & 1 & 0 & 0\\
        1 & T & T^2 & T^3\\
        0 & 1 & 2T & 3T^2
    \end{bmatrix}, \quad
    b_n = \begin{bmatrix}
        n_0 \\
        \dot{s}_0\\
        n_{e, j} \\
        \dot{n}_{e, j}
    \end{bmatrix}.
\end{equation}
The lateral positions are sampled within the track boundaries,
\begin{equation}
    n_{e,j} \in \left[n_r(s_{e, i}) + \frac{d_w}{2} + \Delta_{\text{safety}}, n_l(s_{e, i}) - \frac{d_w}{2} - \Delta_{\text{safety}}\right],
\end{equation}
considering the vehicle width $d_w$ and safety distance $\Delta_{\text{safety}}$.
The interval is divided into positions near the raceline (with denser sampling and includes the terminal racing position $n_{rl, i} (T)$) and those farther away, with the total number of samples is denoted as $N_n$.
The end lateral velocity is chosen based on the longitudinal velocity, calculated as
\begin{equation}
    \dot{n}_{e,j} = \dot{s}_{e,i} \sin (\chi),
\end{equation}
where $\chi$ is the heading in the local frame. 
The lateral trajectory samples are stored accordingly to the corresponding longitudinal trajectories, resulting in a state trajectory matrix of $S \in \mathbb{R}^{N_s \cdot N_n \times N_T \times 4}$.

%? LSTM: we introduce the prediction module to allow the planning algorithm to see ahead and prolong the valid period of the trajectory. [Integrating Algorithmic Sampling-Based Motion Planning]

%? local vs global:
%? The global part is often called the mission, in which the long-term vehicle trajectory is decided. 
%? In racing scenarios, this long-term trajectory is optimally calculated based on the reconstructed map of the known racetrack. 
%? Local/online planners, on the other hand, are what make these applications critical as unexpected obstacles and other dangerous situations might compromise successful navigation [GPU implementation of the Frenet Path Planner]