% !TEX root = ../main.tex
\chapter{Experiments}
\label{chapter:Experiments}

The following sections describe the experiments which were conducted to access the degree of success that the optimization implementation and the spatial sampling planner has had
in comparison to the previous implementation.

%* Include testing platform and simulator explanation
%* Include objectives
%* Include generalized training done

\section{Optimization}
\label{section:Optimization}

This section describes the experiments that were performed to measure the degree of optimization that was accomplished.
The wall collision loss was vectorized to take advantage of SIMD, advantage of multiple register calculation.
This also covers the optimization of the sampling algorithm of DiffPlan.
Two major metric will be looked at; first the full runtime of the algorithm will be assessed, secondly the planning runtime will be measured as the number of sampled trajectory increases.
The initial subsection \ref{subsection:Runtime}, will focus measuring runtime of the four algorithms presented previously and compare them as they were ran on AMD Ryzen 9 5950X 16-Core Processor with SMT,
64 GB of DDR4 RAM, and an NVIDIA GeForce RTX 3090 GPU with 24GB of GDDR6X VRAM.
Additional to the runtime, the scalability of the four algorithms will be assessed as the number of sampled trajectories increase.

\subsection{Runtime}
\label{subsection:Runtime}

The aim is to compare the runtime of the original algorithm, vectorized, gpu focused, and the spatial sampling planner.
This is done by taking a profile trace of the algorithm running using the PyTorch profiler.
The profiler records all traversed functions that it encounters and the length of time each function took as the Python code is ran and saves it as a trace json file to be read later.
The profiling was ran consistently at the same start and end points over the four algorithms as not to contaminate any comparisons.
The experiment was ran five times to prevent any outlier data from overbearing the weight of the measurement, and the runs were averaged to get a final time for each section.
% We aim to see a significant decrease in the runtime. The future work will be presented to make gpu focused better

\subsection{Scaling}
\label{subsection:Scaling}

The aim is to compare the time taken to compute an $n$ number of trajectories vs the log of $n$ to see how it scales as the number of trajectories increase.
Also used PyTorch profiler.

\section{Local Sampling Planner}
\label{section:DiffPlanAndDiffSpatialPlan}

This section covers the experiments performed on the simulator and in real life pertaining to the differences between the DiffPlan and SpatialDiffPlan algorithms.
ROS2 was utilized to run the car for both sim and real.
The experiments utilized real and sim runs to draw some conclusions about the individual runs as well as a combined sim vs real comparison.

FTM Oval track had the same conditions used for both, this was done to draw some insights into the sim2real gap that must be overcome.
Two other tracks were tested to visualize the behavior and ability for each algorithm to show case their local planning ability, path tracking towards the global goal, and finally their lap time improvements.

\subsection{Simulation Runs}

\subsubsection{FTM Oval}


\subsubsection{Austria Track}

\subsubsection{Yas Marina Track}

\subsection{Real Runs}


\subsubsection{FTM Oval}

\subsubsection{Austria Track}

\subsubsection{Yas Marina Track}